Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Subjective Wellbeing

### 2019-10-21 Dissonance_Santiago_v1.tex
Worked on Analysis and Results, included Cochran-Mantel-Haenszel Chi-Squared Test for 3-way tables

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do Drivers Dream of Walking? An Investigation of Dissonance in the Primary Mode of Travel from the Perspective of Subjective Wellbeing},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Do Drivers Dream of Walking? An Investigation of Dissonance in the
Primary Mode of Travel from the Perspective of Subjective Wellbeing}
    \author[Some University]{Author 1\corref{c1}}
   \ead{a1@university.com} 
   \cortext[c1]{Corresponding Author}
    \author[Some School]{Author 2}
   \ead{a2@school.com}

@@

  \begin{abstract}
  Transportation in most of the world has been dominated for decades by a
  fascination with the automobile. Increasingly, however, it is recognized
  that to achieve a variety of economic, environmental, and public health
  policy goals it is important to attract and retain users of public
  transport and active travel. A challenge with the way transportation
  policy is developed is that it tends to focus on cost-benefit and
  utilitarian considerations, which may miss other potential benefits of
  transportation. For this reason, subjective wellbeing (SWB) has been
  proposed as a way to enhance our understanding of the preferences and
  choices of travellers, as well as a way to evaluate the benefits of
  transportation beyond utilitarian considerations. The objective of this
  paper is to analyze the modes that people commonly use, and to what
  extent they are aligned (or not) with a variety of affective and
  utilitarian values. In other words, we are interested in the potential
  for dissonance with respect to the primary mode of travel, from the
  perspective of affective and utilitarian values. The study is based on
  data collected from a sample of travellers in the city of Santiago, in
  Chile. Participants in the study were asked about their usual mode of
  travel, and then were asked to name the mode or modes that they
  associated to affective values (freedom, enjoyment, and happiness) and
  to utilitarian values (safety, time waste, and efficiency). The results
  indicate that users of public transportation experience the most
  dissonance in terms of affective values, and active travellers the
  least. For those travellers who experience dissonance, active travel is
  the mode most commonly associated with affective values.

  To properly assess transportation policy This poses a number of
  challenges Broadening the commonly used transport project assessment
  will be relevant for achieving this goal. While satisfaction with travel
  has been mainly used in the context of daily trips and mainly under
  cost-benefit and utilitarian considerations, subjective wellbeing (SWB)
  over time has risen as an alternative for understanding the preferences
  and choices of transport users. The aim of this paper is to analyse the
  factors that can potentially affect the levels of SWB in the wide
  spectrum of public, private and active transport modes in the context of
  Santiago de Chile. The paper aims to reflect on the possibilities of
  increasing the SWB with a view to enhance the attractiveness of public
  transport and active travel. The paper investigates the affective and
  instrumental/utilitarian factors of SWB, according to modal choice. It
  uses an analysis of correspondence that considers the actual primary
  mode used, versus the mode that each individual associated to both
  utilitarian and affective SWB factors. The results show that there is
  more persistent modal dissonance between positive emotional and
  utilitarian associations within the more disadvantaged groups --
  particularly among users of public transport. The study shows that apart
  from time-savings and efficiency, security has also been
  cross-sectionally sensitive according to the data analysed. Factors such
  as happiness, enjoyment and freedom require far more attention in
  transport planning as they are strongly related to the dimensions of
  subjective and personal wellbeing. Targeting these factors will be
  relevant for attracting new users to public transportation and active
  travel.\\
  \end{abstract}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Transportation planning for decades has focused on providing mobility
for the private automobile. This is a model of development that was
initially introduced in North America as a solution to urbanizations
problems, and that was eventually copied elsewhere (Brown et al., 2009;
Angotti, 1996) . Despite the initial promise of automotive technology,
it is now evident that mobility centred on the private automobile has
given rise to a litany of maladies that are in urgent need of
correction. This includes environmental concerns (i.e., climate change;
Chapman, 2007) as well as other social (Boschmann and Kwan, 2008),
health (Khreis et al., 2016), and equity issues (Bocarejo and Oviedo,
2012). As the real impacts of our societal dependence on the private
autombobile have become increasingly evident, the transportation agenda
has in recent years aimed to shift focus to the reduction of car-use and
towards the creation of mobility polycultures that offer a broader menu
of transportation alternatives than just the private automobile (Miller,
2011) . In order to successfully achieve this goal, it is essential not
only to provide the services and facilities that support public
transportation and active travel, but also to attract new users to these
modes of transportation (Ettema et al.~2011). Within this context, it
has been argued that sustainable transportation policies require all
participants in the transportation system to challenge what Gossling and
Cohen (2014) term ``transportation taboos'': deep-seated ideas
concerning the contribution to emissions by individuals, the inequality
of market-based approaches, and the social and psychological functions
of transportation. With respect to the latter, moving beyond a purely
utilitarian focus, it is important to understand how users perceive
affective and instrumental elements on their transportation experience,
since this can help to incorporate adequate improvements in alternative
transportation, including transit and active modes, to make them more
attractive to users (Anable and Gatersleben, 2005; Domarchi et al.,
2008; Gatersleben and Uzzell, 2007; Steg 2005 ).

A useful lens to investigate the social and psychological functions of
transportation is by means of the concept of subjective wellbeing (SWB.
See, inter alia, De Vos et al., 2013 and Chatterjee et al., 2019) . SWB
is defined by the OECD as ``{[}g{]}ood mental states, including all of
the various evaluations, positive and negative, that people make of
their lives, and the affective reactions of people to their
experiences'' (OECD, 2013, p.~10). Accordingly, the aim of this paper is
to analyse some of the factors that can potentially affect the levels of
SWB in the wide spectrum of public, private and active transport modes
in the context of Santiago de Chile. Previous research suggests that
public transport users experience a negative gap in desired and actual
travel time, whereas active travellers don't seem to mind somewhat
longer trips than what strictly needed (e.g., Páez and Whalen, 2010). In
the context of a Latin American country, historically the poor travel
experience of public transit users and active travellers reflects the
social inequalities of social groups that have no alternative but to
move in these types of modes while living far from the work centres and
main activities. The paper aims to reflect on the possibilities of
increasing the SWB while travelling, and how this could increase the
attractiveness of public transport and active travel in the
socio-demographic groups studied. This can help not only to reduce the
use of private transport but also decrease the inequality gap between
these users.

The paper contributes to the literature in three areas. First, it
contributes to an emerging literature on the topic in the context of the
Global South (Bejarano et al., 2017 ; Shao and Liang, 2019 ; Van et al.,
2014 ; Zorrilla et al., 2019); to the best of our knowledge, the case of
Chile has not yet been reported. Second, although there is literature on
the enjoyment of commute and other emotional factors, from a hedonic and
even eudaimonic perspectives, the analysis has yet to be applied more
fully in terms of distributional issues -- i.e.~which groups are mostly
disadvantaged. Finally, the analysis shows the attitudes of people
towards their primary mode and their perception towards `ideal modes' --
implying their preferences, even in situations when their ideal mode is
not part of their actual choice set. The attitudes of travellers towards
transport modes are critical factors to be considered by policy-makers
in case they want to promote and increase the use of public transport or
active modes (e.g., Bejarano et al., 2019). The paper explores these
positive and negative associations within the transport options used in
Santiago.

\hypertarget{background}{%
\section{Background}\label{background}}

The transportation community has for the most part already agreed to the
need of looking at mobility and transport issues from the lens of
wellbeing, to broaden the commonly used utilitarian perspective to
evaluate projects. Users' travel satisfaction (Bergstad, 2001) has been
the primary way to direct this challenge towards everyday mobility,
resulting on research about satisfaction with different transport modes.
For example, numerous studies report that car users have a higher level
of satisfaction compared to other transport modes (Redmond and
Mokhtarian, 2001 ; Bergstad et al.~2011; Eriksson et al.~2013; Whalen et
al., 2013). In a similar way, there are multiple reports that active
travel also tends to yield high levels of satisfaction (Smith, 2017;
St-Louis et al.~2014; Paez and Whalen, 2010; Whalen et al., 2013), In
contrast, public transport users tend to assess their experience more
negatively (De Vos et al.~2016; Abenoza et al.~2017; Ettema et al.~2012;
Páez and Whalen, 2010). Multi-modal trips also influence satisfaction
levels; for instance, when an individual chooses a particular mode of
transportation, she will report a higher level of satisfaction with that
chosen mode -- perhaps as a form of post hoc validation (Susilo and
Cats, 2014).

While the use of travel satisfaction has been mainly used in the context
of daily trips -- typically been linked to cost-benefit and utilitarian
measurements --, the evaluation of Subjective Wellbeing (SWB) over time
has risen as an alternative measure. In the field of travel behaviour,
Ettema et al.~(2010; p.~725) have defined SWB as the degree to which an
individual positively evaluates the overall quality of their lives,
where the general life satisfaction encompasses a more extended
temporality -- which implies assuming a tendency to be more stable over
time. This concept has prompted a growing literature that complements
and applies SWB in a broader range of satisfaction scales and
situations. The definition of other factors such as travel choice mode,
attitudes and external elements of the built environment has been
studied for a broader understanding of the changes produced in the SWB.
As these factors do not necessarily apply to the general life
satisfaction on the long term, the studies have aimed to determine both
the direct and indirect effects on the perception of users (see, e.g.~Ye
and Titheridge 2017). Other concepts have also emerged as the
Satisfaction with Travel Scale (STS), a measurement devised by Ettema et
al.~(2011) as well as different scales based on people's travel
perceptions. De Vos et al.~(2015) for instance, explore in detail the
underlying dimensions of the affective domain of STS in which SWB is
based (for more on STS see also Friman et al.~2013).

Recent literature on SWB and its link with transport have demonstrated a
relationship between people's perceptions and satisfaction with their
daily travel (Smith, 2017; St-Louis et al.~2014; Paez and Whalen, 2010;
Whalen et al., 2013). Scholars have shown that accessibility has been
the most developed factor that influences people's wellbeing (Delbosc,
2012), and activities have a direct impact on travel satisfaction
(Bergstad et al., 2010). Delbosc (2012; 28) for instance, has summarised
the most significant influences on psychological wellbeing: poverty and
employment, meaningful relationships and health. However, understanding
the components affecting people's perceptions implies the
differentiation between the emotional (also named as symbolic-affective)
and the instrumental reasons (Bergstad, 2001). Steg et al.~(2011) have
compared symbolic-affective opposed to instrumental-reasoned motives
based on car-use, and other studies have also found associations between
affective and symbolic aspects of car-use (see, e.g.~Gatersleben, 2007;
Lois \& López-Sáez, 2009). Previous studies have demonstrated how
socio-demographic factors affect the levels of SWB. The effect of income
on SWB (Clarck and Oswald, 1996, Ferrer-i-Carbonell, 2005); education
and unemployment (Argyle, 1999); age (Diener and Suh, 1997) and gender
(Tesch-Römer et al., 2008) have already been studied. Recent research
also suggests the links between commuting, SWB and emotional wellbeing
assessment (Olson et al.~2013; Kahneman et al.~2004). However, more
research is needed to understand how these socio-demographic variables,
connect as well with the decisions on main transport mode choices; or
research distinguishing between hedonic (journey experience) and
eudaimonic (flourishing) value of trips.

The research needs already recognized in the developed world are also
markedly acute in the context of the Global South, where historical
inequality has tended to create a symbolic attachment to the automobile,
in addition to negative connotations for public transport and active
travel (Zorrilla et al., 2019). In this way, there is an emerging
literature that investigates affective factors in travel behaviour in a
number of developing countries. A cross-country study in Asia revealed
that the affective factors of public transportation and car use are
important, and in particular the social orderliness of transit was
suggested as a way to make this mode more attractive to users (Van et
al., 2014). In terms of active travel, a study in China found that
attitudes that embrace new styles and technologies despite their cost
are associated with the intention to continue using shared bicycles
(Shao and Liang, 2019). The importance of affective factors for policy
and planning is further highlighted by research in Colombia that shows
how users felt proud using a bicycle shared system, in addition to
experiencing feelings of belonging to a civic culture and the enjoyment
and pleasure of travel itself (Bejarano et al., 2019). This paper
contributes to further our understanding of affective and instrumental
factors in travel behaviour in a developing country.

\hypertarget{data-and-methods}{%
\section{Data and Methods}\label{data-and-methods}}

This paper implements an analysis of correspondence comparing the
primary mode of transportation with affective and utilitarian SWB
factors in a sample of travellers in Santiago de Chile.

\hypertarget{sample}{%
\subsection{Sample}\label{sample}}

The study is based on a survey conducted in Santiago during the months
of November and December 2016, that collected information on a wide
range of travel and related issues. The data collection considered a
quota-sampling method for gathering the information, considering the
socio-demographic information from Pre-Census of 2012. An equal
representation of both genders and a representation of the proportion of
inhabitants per area have been chosen as relevant characteristics of the
sample. In total, 451 persons validly completed the survey. In the first
part of the questionnaire the participants were asked about basic
socio-demographic data and main transport mode for commuting. In the
second part, the questionnaire asked about the modes that people
associate with affective and instrumental positive concepts of
commuting, as well as the levels of overall satisfaction with the trips.
These two sections of the survey have been considered for the analysis
of this paper. The second part of the survey asked about the association
between the primary transport mode and 18 positive and negative
feelings. The question is: ``Indicate the modes that you associate to
the following feelings''. The respondents could indicate all the modes
they associated to the feelings on the list. Three affective and three
instrumental/utilitarian factors, all positive, have been considered for
this analysis: freedom, enjoyment, happiness (affective), and security,
time savings, efficiency (instrumental/utilitarian). The questionnaire
provides a binary indicator that describes the presence or absence of
modal correspondence for each one the selected six factors. The analysis
has considered socioeconomic variables (gender, age, education,
occupation, income, disabilities), as well as other variables as driving
licence, journey times, monthly expenditure on transport and
satisfaction with the overall trip.

{[}REPORT DESCRIPTIVE STATISTICS? Include definitions, e.g., is K-12
equivalent to ``minimal mandatory education''? What is Tech Univ? What
are the range of values of low, middle, high income? Journey time is in
minutes{]}

\hypertarget{analytical-approach}{%
\subsection{Analytical Approach}\label{analytical-approach}}

The analytical approach in this paper is based on correspondence.
Previous research has developed the concept of modal dissonance for
referring to the mismatch between the actual mode and the mode that
enables users to experience positive affective or instrumental factors
(De Vos et al., (2012; 2014) and Schwanen and Mokhtarian, 2004, 2005a,
2005b). More concretely, Schwanen and Mokhtarian (2004; 89) define
dissonance based on the incongruence between residential neighbourhood
where the individual currently resides and her preference structure
toward such characteristics of the residential environment. In this
paper we consider correspondence, which happens when travellers
associate the positive utilitarian and affective SWB factors to the mode
they have used as their primary transport mode for commuting or
performing their main activities. The opposite of correspondence would
be modal dissonance.

More specifically, the concept of modal dissonance means that a user of
mode X identifies a different mode Y when associating affective values.
For example, a user of private transport (i.e., mode ``Car'') may select
mode ``Bicycle'' when asked about the of the association between the
concept of freedom to a transport mode. In this case, the person who
uses the mode ``Car'' as their primary mode of transportation attaches
positive emotional and instrumental factors not to that mode, but to a
different one -- in this case to the bicycle. It could be the case that
the modal dissonance corresponds just to one of the types of factors,
either affective or instrumental. For example, it is possible that a
person who uses the Metro as their main transport mode associates that
mode to the instrumental factors (security, time savings, efficiency) --
but relates positive emotional factors (freedom, enjoyment, happiness)
to a different mode, say to the bicycle.

The ways for measuring equity in transport projects and the most
vulnerable groups affected by transport have been the aspects that have
nurtured the focus of this study. The hypothesis is that public
transport users will show a lower correspondence between their primary
mode and the modes they relate with positive factors of SWB. This is
relevant as more disadvantaged people have less options of choosing
their preferred transport modes and they tend to associate positive
affective and instrumental factors to modes that they are not able to
use. These perceptions evidence equity gaps between groups -- where the
most vulnerable groups are much more affected. The methods used are
analysis of correspondence and multinomial regression, both around the
associations of transport modes and affective and instrumental factors,
derived from the SWB framework.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{who-experiences-dissonance}{%
\subsection{Who experiences
dissonance?}\label{who-experiences-dissonance}}

The first step of the analysis seeks to understand whether there are
statistical associations between the variables of interest. The analysis
considers the cases where there is a lack of correspondence between the
primary mode choice and the 6 SWB/instrumental factors previously
described -- or in other words, where there is modal dissonance.

Table 1 shows the results of the analysis of correspondence and the
dissonances between the primary mode of travel on the one hand, and SWB
and instrumental factors on the other. As seen in the table, there are
five categories that we use for the analysis. Three of these categories
are socio-economic and demographic, namely age, level of education,
income. The other two categories are transportation related, namely
primary mode of travel and journey time. Correspondence is tested in
every case by means of the \(\chi^2\) statistic. It is interesting to
note that the only category for which all affective/instrumental factors
are significant at better than 5\% level of confidence is mode of
travel.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Subjective Wellbeing},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance
from the Perspective of Subjective Wellbeing}
    \author[Some University]{Author 1\corref{c1}}
   \ead{a1@university.com} 
   \cortext[c1]{Corresponding Author}
    \author[Some School]{Author 2}
   \ead{a2@school.com}

@@

  \begin{abstract}
  Transportation in most of the world has been dominated for decades by a
  fascination with the automobile. Nonetheless, there is an increasing
  recognition that to achieve a variety of economic, environmental, and
  public health policy goals it is important to attract and retain active
  travelers and users of public transportation. A challenge with the way
  transportation policy is developed, however, is that it tends to focus
  on utilitarian considerations, which may miss other potential aspects of
  transportation that users value. In particular, subjective wellbeing
  (SWB) has been proposed as a way to enhance our understanding of the
  preferences and choices of travellers, as well as a way to evaluate the
  benefits of transportation beyond utilitarian considerations. The
  objective of this paper is to analyze the modes that people commonly
  use, and to what extent they are aligned (or not) with a variety of
  affective values associated with SWB. In other words, we are interested
  in the potential for dissonance with respect to the primary mode of
  travel, from the perspective of affective values. The study is based on
  data collected from a sample of travellers in the city of Santiago, in
  Chile. Participants in the study were asked about their usual mode of
  travel, and then were asked to name the mode or modes that they
  associate with the affective values of freedom, enjoyment, happiness,
  poverty, luxury and status. The results indicate that users of public
  transportation experience the most dissonance in terms of affective
  values, and active travellers the least. For those travellers who
  experience dissonance, active travel is the mode most commonly
  associated with freedom, enjoyment, and happiness, public transportation
  is most commonly associated with poverty, and the automobil is most
  commonly associated with luxury and status.
  \end{abstract}

@@

\hypertarget{introduction}{%
\section{1. Introduction}\label{introduction}}

Transportation planning for decades has focused on providing mobility
for the private automobile. This is a model of development that was
initially introduced in North America as a solution to problems created
by rapid urbanization, and that was eventually copied elsewhere
(Angotti, 1996; Brown et al., 2009). Despite the initial promise of
automotive technology, it is now evident that mobility centred on the
private automobile has given rise to a litany of maladies that are in
urgent need of correction. This includes environmental concerns (i.e.,
climate change; Chapman, 2007) as well as numerous other social
(Boschmann and Kwan, 2008; Lucas, 2019, 2012), health (Khreis et al.,
2016; Milne, 2012), and equity issues (Bocarejo and Oviedo, 2012;
Martens et al., 2012; Pereira et al., 2017).

As the real impacts of our societal dependence on the private
autombobile have become increasingly evident, the transportation agenda
has aimed to shift focus to the reduction of car use and towards the
creation of mobility polycultures that offer a broader menu of
transportation alternatives than primarily (or even just) the private
automobile (Miller, 2011). In order to successfully achieve this goal,
it is essential not only to provide the services and facilities that
support public transportation and active travel, but also to attract new
users to these modes of transportation (Ettema et al., 2011). Within
this context, it has been argued that sustainable transportation
policies require all participants in the transportation system to
challenge what Gossling and Cohen (2014) termed \emph{transportation
taboos}: deep-seated ideas concerning the contribution to emissions by
individuals, the inequality of market-based approaches, and the social
and psychological functions of transportation. With respect to the
latter, it is important to move beyond a purely utilitarian focus if we
are to understand how the affective value of the transportation
experience can be leveraged to improve alternative transportation -
particularly transit and active modes (Anable and Gatersleben, 2005;
Domarchi et al., 2008; Gatersleben and Uzzell, 2007; Steg, 2005).

A useful lens to investigate the social and psychological functions of
transportation is by means of the concept of subjective wellbeing (SWB;
see, inter alia, De Vos et al., 2013; and Chatterjee et al., 2019). SWB
is defined by the OECD as ``{[}g{]}ood mental states, including all of
the various evaluations, positive and negative, that people make of
their lives, and the affective reactions of people to their
experiences'' (\textbf{OECD, 2013, p.~10}). As memorably put by Steg
(2005): is the use of car a must, or is it a lust? A close alignment, or
consonance, between affective values and the mode of transportation used
to travel can result in greater subjective wellbeing and increase the
probability of choosing that mode (i.e., by satisfying the \emph{lust});
on the other hand, dissonance, that is, the lack of concordance between
the mode used and affective values would be detrimental to subjective
wellbeing, especially of those for whom the use of the mode is a
\emph{must} (De Vos, 2019; Mokhtarian and Pendyala, 2018). With these
considerations in mind, one aim of the present research is to
investigate \emph{who} experiences dissonance with respect to their
primary mode of travel and affective values, namely freedom, enjoyment,
happiness, feelings of poverty, luxury, and status. Furthermore, it is
possible, even if use of a particular mode is a \emph{must}, that
travellers may still \emph{lust} for something else - in other words,
the grass may in fact look greener from the window of the car. For this
reason, we also aim to investigate which mode or modes are most commonly
associated with affective values, with a focus on those travellers who
experience dissonance in their primary mode.

Research is based on data collected from a sample of travellers in the
city of Santiago in Chile. Survey respondents were asked about their
primary mode of travel, and also about the mode or modes that they
associate with the affective values mentioned above. The paper
contributes to the literature in three ways, as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The research reported here contributes to an emerging literature on
  the topic of transportation and affective values in the context of the
  Global South (Al-Ayyash and Abou-Zeid, 2019; Bejarano et al., 2017;
  Shao and Liang, 2019; Van et al., 2014; Zorrilla et al., 2019); to the
  best of our knowledge, the case of Chile has not yet been reported.
\item
  Although there is an extensive literature on the enjoyment of commute
  and other affective values (see for instance Paez and Whalen, 2010;
  Redmond and Mokhtarian, 2001; Whalen et al., 2013; Ye and Titheridge,
  2017), from a hedonic and even eudaimonic perspectives the analysis
  has yet to be applied more fully in terms of distributional issues --
  i.e.~which groups more commonly experience dissonance (see De Vos,
  2018).
\item
  The analysis shows the attitudes of people towards their primary mode
  and their perception towards `ideal modes' -- implying their
  preferences, even in situations when their ideal mode is not part of
  their actual choice set. More concretely, the results indicate that
  users of public transportation experience the most dissonance in terms
  of affective values, and active travellers the least. For those
  travellers who experience dissonance, active travel is the mode most
  commonly associated with freedom, enjoyment, and happiness, public
  transportation is most commonly associated with poverty, and the
  automobil is most commonly associated with luxury and status. We also
  find that there are some substantial variations in dissonance by age,
  education, income, and typical commute time. The attitudes of
  travellers towards transport modes are critical factors to be
  considered by policy-makers in case they want to promote and increase
  the use of public transport or active modes (Bornioli et al., 2019; De
  Vos et al., 2019; De Vos and Witlox, 2017; Garling et al., 2019;
  Redman et al., 2013).
\item ~
  \hypertarget{background}{%
  \section{Background}\label{background}}
\end{enumerate}

A consensus has emerged in the transportation community regarding the
need to complement the traditional utilitarian perspective of
transportation by looking at mobility and transport issues from the lens
of their affective functions. The affective value of transportation in
turn is important due to its potential to improve or detract from SWB.
One of the primary ways to explore this has been the satisfaction that
travelers feel towards their every day mobility experience (e.g.,
Cecilia Jakobsson Bergstad et al., 2011). As a consequence, there is a
wealth of research on satisfaction with the use of different modes of
transportation. For example, numerous studies report that car users
often have a higher level of satisfaction compared to other transport
modes (C. J. Bergstad et al., 2011; Eriksson et al., 2013; Redmond and
Mokhtarian, 2001; Whalen et al., 2013; but see Handy and Thigpen, 2019).
In a similar way, there are multiple reports that active travel also
tends to yield high levels of satisfaction (Gatersleben and Uzzell,
2007; Handy and Thigpen, 2019; Paez and Whalen, 2010; Smith, 2017;
St-Louis et al., 2014; Whalen et al., 2013), In contrast, public
transport users tend to assess their experience more negatively (Abenoza
et al., 2017; De Vos et al., 2016; Gatersleben and Uzzell, 2007; Handy
and Thigpen, 2019; Paez and Whalen, 2010). Multi-modal trips also
influence satisfaction levels; for instance, when an individual chooses
a particular mode of transportation, she will report a higher level of
satisfaction with that chosen mode -- perhaps as a form of \emph{post
hoc} validation (Susilo and Cats, 2014).

While the use of travel satisfaction has been mainly used in the context
of daily trips -- typically linked to cost-benefit and utilitarian
measurements --, the evaluation of Subjective Wellbeing (SWB) over time
has risen as an alternative measure. In the field of travel behaviour,
Ettema et al.~(2010, p. 725) define SWB as the degree to which an
individual positively evaluates the overall quality of their lives,
where the general life satisfaction encompasses a more extended
temporality -- which implies assuming a tendency to be more stable over
time. This concept has prompted a growing literature that complements
and applies SWB in a broader range of satisfaction scales and
situations. The definition of other factors such as travel choice mode,
attitudes and external elements of the built environment have been
studied for a broader understanding of the changes produced in the SWB
(e.g., Handy and Thigpen, 2019). As these factors do not necessarily
apply to the general life satisfaction on the long term, the studies
have aimed to determine both the direct and indirect effects on the
perception of users (see, e.g., Ye and Titheridge, 2017). Other concepts
have also emerged as the Satisfaction with Travel Scale (STS), a
measurement devised by Ettema et al.~(2011), as well as different scales
based on people's travel perceptions. De Vos et al.~(2015), for
instance, explore in detail the underlying dimensions of the affective
domain of STS on which SWB is based (for more on STS see also Friman et
al., 2013).

Recent literature on SWB and its link with transport have demonstrated a
relationship between people's perceptions and satisfaction with their
daily travel (e.g., Smith, 2017; Mokhtarian and Pendyala, 2018; St-Louis
et al., 2014). Scholars have shown that accessibility has been the most
developed factor that influences people's wellbeing (Delbosc, 2012), and
activities have a direct impact on travel satisfaction (Cecilia
Jakobsson Bergstad et al., 2011). Delbosc (2012, p. 28), for instance,
has summarised the most significant influences on psychological
wellbeing: poverty and employment, meaningful relationships and health.
However, understanding the components affecting people's perceptions
implies the differentiation between affective (also named as
symbolic-affective) and instrumental values (C. J. Bergstad et al.,
2011). Steg et al.~(2011) have compared symbolic-affective opposed to
instrumental-reasoned motives based on car-use, and other studies have
also found associations between affective and symbolic aspects of
car-use (see, e.g., Gatersleben and Uzzell, 2007; Lois and Lopez-Saez,
2009). Previous studies have also demonstrated how socio-demographic
factors affect the levels of SWB. The effect of income on SWB (Clark and
Oswald, 1996; Ferrer-i-Carbonell, 2005); education and unemployment
(Argyle et al., 1999); age (Diener and Eunkook Suh, 1997), and gender
(Tesch-Römer et al., 2008) have been extensively studied. Recent
research also suggests the links between commuting, SWB and emotional
wellbeing assessment (\textbf{Olson et al.~2013}; \textbf{Kahneman et
al.~2004}). However, more research is needed to understand how these
socio-demographic variables connect as well with the affective responses
to mode of travel (St-Louis et al., 2014).

The research needs already recognized in the developed world are also
markedly acute in the context of the Global South, where historical
inequality has tended to create a symbolic attachment to the automobile,
in addition to negative connotations for public transport and active
travel (Zorrilla et al., 2019). In this way, there is an emerging
literature that investigates affective factors in travel behaviour in a
number of developing countries. A cross-country study in Asia revealed
that the affective factors of public transportation and car use are
important, and in particular the social orderliness of transit was
suggested as a way to make this mode more attractive to users (Van et
al., 2014). In terms of active travel, a study in China found that
attitudes that embrace new styles and technologies despite their cost
are associated with the intention to continue using shared bicycles
(Shao and Liang, 2019). The importance of affective factors for policy
and planning is further highlighted by research in Colombia that shows
how users felt proud using a bicycle shared system, in addition to
experiencing feelings of belonging to a civic culture and the enjoyment
and pleasure of travel itself (Bejarano et al., 2017). This paper
contributes to further our understanding of affective values in travel
behaviour in a developing country.

\hypertarget{data-and-methods}{%
\section{3. Data and Methods}\label{data-and-methods}}

\hypertarget{context}{%
\section{3.1 Context}\label{context}}

Words to describe Santiago go here.

\hypertarget{sample}{%
\section{3.2 Sample}\label{sample}}

The study is based on a survey conducted in the city of Santiago during
the months of November and December 2016, that is, the end of the Spring
and beginning of Summer. The survey collected information on a wide
range of travel and related issues. The data collection considered a
quota-sampling method for gathering the information, considering the
socio-demographic information from Pre-Census of 2012. An equal
representation of both genders and a representation of the proportion of
inhabitants per area were chosen as relevant characteristics of the
sample. In total, there were \(n=451\) valid surveys, although not every
survey was complete and there were missing responses for some answers.

The survey was structured in eight sections, as follows: 1) Individual
characteristics of respondent; 2) Health; 3) Feelings and emotions; 4)
Reasons for travel and planning travel; 5) Social interaction; 6) Nature
and sustainability; 7) Information and telecommunications; 8) Built
environment; and 9) Commuting. For the present study, we draw data from
sections 1), 3) and 9). In terms of individual characteristics of the
respondents and their commute, participants were asked about basic
socio-demographic information, including age, level of education,
income, and the typical duration of their regular commute. The
descriptive statistics of the sample appear in Figure
\ref{fig:descriptive-statistics}. The sample tends to be younger, and
well-educated, with an almost uniform distribution of income levels. The
trend in typical commute time is towards longer commutes.

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v1_files/figure-latex/plot-descriptive-statistics-1.pdf}
\caption{\label{fig:descriptive-statistics}Descriptive statistics of the
sample}
\end{figure}

In addition, respondents were asked about their primary mode of travel
for their regular commute. The modes available were Car, Taxi, Colectivo
(a form of shared ride, intermediate in flexibility and capacity between
taxi and bus); Motorcycle; Metro; Bus; Bicycle; Walking. As seen in the
top panel of Figure \ref{fig:primary-mode-travel}, the three most common
modes of travel are Metro, Bus, and Car, followed by Walking and Biycle.
For the analysis, we aggregate these modes into the following categories
(bottom panel of Figure \ref{fig:primary-mode-travel}): Car, Active
(Walking + Bicycle), Public (Metro + Bus), and Other (Taxi + Colectivo +
Motorcycle).

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v1_files/figure-latex/figure-primary-mode-travel-1.pdf}
\caption{\label{fig:primary-mode-travel}Frequency of primary mode used
for regular commute; top panel: all modes, bottom panel: aggregated
modes}
\end{figure}

Of particular interest for the present study is the following question
in Part 3) of the survey:

\begin{quote}
\begin{quote}
Q: Please indicate the mode(s) of transport that you relate to the
following feelings and concepts
\end{quote}
\end{quote}
